\speculake: Hiding Malware in Speculative Execution

### 2018-08-02 paper/architecture.tex
arch pass

We further assume that the malware has a specific trigger that the analyst is
not privy to. In our examples, this trigger is often behavior exhibited by some
other (potentially benign) process running on the same system as the malware.
As the adversary is able to control when such a trigger is deployed (potentially
remotely), the analyst will not be able to observe or force this trigger to
happen at will. We emphasize that while this may also be true for existing
trigger-based malware, analysts can often reverse engineer the trigger out of
the malware, for example by observing control flow within the malware and using
adaptive fuzzers~\cite{afl} to generate inputs that explore other execution
paths of the binary. In contrast, \speculake malware's trigger depends on
behavior in seemingly unrelated binaries, influencing malware behavior through
channels invisible to the analyst. As with typical malware, the analyst may
attempt to reverse engineer the trigger to revela the malware's behavior, but we
will show (in Section~\ref{subsec:nested-spec}) how to defeat this type of
analysis.

@@

In \speculake, we cause the CPU to mis-speculate the destination of an indirect
branch in the payload program, causing it to speculatively execute 
instructions that are never truly executed. We term this speculative destination
the \emph{speculative entry point}. \speculake uses indirect jumps to allow
speculative execution from \emph{any} instruction in the payload process'
address space. Because it can jump to any instruction, the malware analyst has a
difficult task in determining where a payload program's speculative entry point
is.

@@

We observe that Intel CPUs consider three types of x86\_64 indirect branches:
\texttt{retq}, \texttt{callq *\%rax}, and \texttt{jmpq *\%rax}\footnote{other
general purpose registers besides \texttt{\%rax} can be used as well}.
We created a simple trigger program that performs a series of 28 indirect
branches
using \texttt{jmpq *\%rax} instructions. Between each jump, we increment
\texttt{\%rax} to account for the number of bytes between jumps. For the last
jump, we load a function pointer into \texttt{\%rax} and do a final indirect
branch using \texttt{callq *\%rax}.

In our payload program, we first perform the same 28 indirect jumps. We ensure
the source address of these jumps is the same as in the trigger program by
manually defining their containing function at a fixed address inside a linker
script. We also do the final indirect call to a function pointer, but with two
differences. First, the destination of the function pointer is a different
address, and second, the location of the function pointer in memory is uncached.
This forces the CPU to predict the destination of the indirect call while it
waits for the function pointer to load from memory. Due to the similar
history of branches with the trigger program,
the CPU will (incorrectly) predict the destination to be the same as the one in
the trigger program, which determines the speculative entry point for the
payload. Even though the in-order execution of
payload program never executes or even reads from this address, the CPU will
briefly execute instructions there speculatively.

Eventually, the dereference of the uncached function pointer in payload program
will be resolved, and the CPU will recognize it has incorrectly predicted the
destination of its \texttt{callq}/\texttt{jmpq}/\texttt{retq} instruction. The
results from the speculative entry point instructions will be discarded, and the
CPU will continue executing from the correct destination. However, as the
speculative code changes what is loaded into the cache based on its results, it
can covertly communicate its results to the ``real world'' program.

@@

We performed several experiments to determine how much computation can be
performed
speculatively, as well as what components are responsible for the limit.
We report results from our experiments on an Intel~Xeon-1270,
though we note we found similar results across other Intel processor generations,
including an i5-7200U, % jack
an i5-4300U, % ewust laptop
and an i5-4590. % ewust desktop
% For simplicity, we report results from only the Xeon-1270.

@@

We also tested whether instruction complexity or data dependencies impact the
number of instructions that can be completed. We find that data dependencies and
instruction complexity both have an impact on the number of instructions that
can be executed. Instruction complexity is determined by the number of $\mu$ ops
that the instruction uses, which appears to be what is tracked in the ROB. For
instance, on our architecture, the 64-bit \texttt{idiv} instruction takes 59
$\mu$ ops, and we can execute up to 3 of them in the speculative world.
Meanwhile, we can execute up to 18 32-bit \texttt{idiv} instructions, which each
take 9 $\mu$ ops~\cite{intel-instruction-tables}.

@@

Most notably instructions that use the extended x86 registers are still valid
within the speculative context. Specifically, Intel's hardware accelerated
AES-NI encryption and decryption instructions, which each use 128-bit registers.
As shown in Figure~\ref{fig:spec-capacity}, speculative environments can
complete a significant number of AES rounds---more than enough to decrypt a
full block using AES-CBC. We investigate the use of AES instructions in the
spculative environment further in Sections~\ref{subsec:decryption}.

@@

We find that when executing speculatively, the number of instrutions completed
is rarely limited by cache miss duration. Instead
ROB size and processor mechanisms for resolving
data dependencies define an upper boundary of approximately 150
instructions\footnote{While \texttt{nop} is able to execute up to the full 220
ROB capacity, instructions that do useful work cannot reach this limit}.

@@

We explore the ability for the CPU to ``double speculate'', where a second
stalled indirect jump while the CPU is already speculating causes it to predict
the target and
speculate a second time. For instance, suppose a payload program truly jumps to
target A, but the CPU is mistrained by a trigger program that jumps to B, thus
causing the payload program to speculatively execute at B. At B, suppose there
is a second indirect jump, perhaps using the same register as the first jump
(which has still not resolved). If the trigger program jumps to C, the payload
program will speculate a second time and continue speculative execution at C.

We find that not all Intel CPUs support nested speculation. For
example, it appears Haswell chips do not speculate while already speculating,
but nonetheless support non-nested \speculake. Both Sandy Bridge (which
preceeded Haswell) and Kaby Lake (which followed Haswell) support nested
speculation. We find that when a CPU does support nested speculation, there
appears to be no limit to nested depth besides the speculative instruction
limit. We use a 16-deep nested speculation in Section~\ref{subsec:impl-aes}
to thoroughly protect speculative decryption keys from reverse
engineering.

@@

We summarize our findings into a \emph{speculative primitive}, which allows us to
speculatively (and covertly) perform on the order of 100 arbitrary
instructions while an accompanying trigger program is running, and communicate
a short (e.g. single byte) result to the real
world via a cache side channel. These speculative instructions are able to read
from any real-world memory or registers, but they cannot perform updates or
writes directly. To update memory, the speculative instructions must communicate
to the real world through a cache side channel.

We note a performance tradeoff between the size of communication (e.g. 4 bits vs 8
bits) and the time it takes the real world to recover the result from the side
channel. Using Flush+Reload~\cite{yarom2014flush+} as our cache side channel,
recovering the result requires accessing all elements in an array exponential in
the size of the result (e.g. $2^8$ array reads to recover an 8-bit result). 
Therefore, there is a
performance advantage for keeping the size of the result small, and communicating
out small pieces of information that are aggregated by the real world.

>>>>

We further assume that the malware has a specific trigger that the analyst is
not privy to. In our examples, this trigger is often behavior exhibited by some
other (potentially benign) process running on the same system as the malware.
As the adversary is able to control when such a trigger is deployed (potentially
remotely), the analyst will not be able to observe or force this trigger to
happen at will. We emphasize that while this may also be true for existing
trigger-based malware, analysts can often reverse engineer the trigger out of
the malware, for example by observing control flow within the malware and using
adaptive fuzzers~\cite{afl,stephens2016driller} to generate inputs that explore other execution
paths of the binary. In contrast, \speculake malware's trigger depends on
behavior in seemingly unrelated binaries, influencing malware behavior through
channels that are effectively invisible to the analyst. As with typical malware, the analyst may
attempt to reverse engineer the trigger to reveal the malware's behavior, but we
will show (in Section~\ref{subsec:nested-spec}) how to defeat this type of
analysis.

@@

In \speculake, we cause the CPU to mis-speculate the destination of an indirect
branch in the payload program, causing it to speculatively execute 
instructions that are never truly executed. We term the destination where speculation begins
the \emph{speculative entry point}. \speculake uses indirect jumps to allow
speculative execution from \emph{any} instruction in the payload process'
address space. Because it can jump to any instruction, the malware analyst has a
difficult task in determining where a payload program's speculative entry point
is.

@@

We observe that Intel CPUs consider three types of x86\_64 indirect branches:
\texttt{retq}, \texttt{callq *\%rax}, and \texttt{jmpq *\%rax}\footnote{other
general purpose registers besides \texttt{\%rax} can be used as well}.
We created a simple trigger program that performs a series of 28 indirect
branches
using \texttt{jmpq *\%rax} instructions, as we found that 28 jumps was the minimum
needed to reliably train the indirect branch predictor. 
Between each jump, we incremented
\texttt{\%rax} accordingly to continue on to the next jump. After these 28
jumps, we load a function pointer into \texttt{\%rax} and do a final indirect
branch using \texttt{callq *\%rax}. In our trigger, we perform these jumps
        repeatedly in a loop.

In our payload program, we first perform the same 28 indirect jumps. We ensure
the source and destination addresses of these jumps is the same as in the trigger program by
manually defining their containing function at a fixed address inside a linker
script. We also do the final indirect call to a function pointer, but with two
differences. First, the destination in the function pointer is a different
address, and second, the memory location of the function pointer itself is uncached.
This forces the CPU to predict the destination of the final indirect call while it
waits for the function pointer to load from memory. Due to the similar
history of branches with the trigger program,
the CPU will (incorrectly) predict the destination to be the same as the one in
the trigger program, which determines the speculative entry point for the
payload. Even though the in-order execution of
payload program never executes or even reads from this address, the CPU will
briefly execute instructions there speculatively.

Eventually, the dereference of the uncached function pointer in the payload program
will be resolved, and the CPU will recognize it has incorrectly predicted the
destination of its \texttt{callq} instruction. The
results from the speculative entry point instructions will be discarded, and the
CPU will continue executing from the correct destination. However, the
speculative code can change what is loaded into the cache based on its
computation, allowing it to covertly communicate its results to the ``real world'' program.

@@

We performed several experiments to determine how much computation can be
performed
speculatively, as well as what components are responsible for the limit.
We report results from our experiments on an Intel~Xeon-1270 (Sandy Bridge),
though we note we found similar results across other Intel processor generations,
including an i5-7200U (Kaby Lake), % jack
an i5-4300U (mobile Haswell), % ewust laptop
an i5-4590 (desktop Haswell), % ewust desktop
as well as an AMD EPYC 7251. % envy
% For simplicity, we report results from only the Xeon-1270.

@@

We also tested whether instruction complexity or data dependencies impact the
number of instructions that can be completed. We find that data dependencies and
instruction complexity both have an impact on the number of instructions that
can be executed. Instruction complexity is determined by the number of $\mu$~ops
that the instruction uses, which appears to be what is tracked in the ROB. For
instance, on our architecture, the 64-bit \texttt{idiv} instruction takes 59~$\mu$~ops,
and we can execute up to 3 of them in the speculative world.
Meanwhile, we can execute up to 18 32-bit \texttt{idiv} instructions, which each
take 9~$\mu$~ops~\cite{intel-instruction-tables}. These suggest we can execute
on the order of 175~$\mu$~ops before the speculative world expires.

@@

Most notably instructions that use the extended x86 registers are still valid
within the speculative context. Specifically, Intel's hardware accelerated
AES-NI encryption and decryption instructions, which each use 128-bit registers.
As shown in Figure~\ref{fig:spec-capacity}, speculative environments can
complete a significant number of AES rounds---over 100 rounds in our
experiments, more than enough to decrypt a
full block using simple AES modes (e.g. AES-CTR). We investigate the use of AES
instructions in the spculative environment further in
Sections~\ref{subsec:decryption}.

@@

We find that when executing speculatively, the number of instrutions completed
is rarely limited by cache miss duration. Instead
ROB size and processor mechanisms for resolving
data dependencies define an upper boundary of approximately 150
instructions\footnote{While \texttt{nop} is able to execute up to the full 220
ROB capacity, instructions that do useful work (and/or use multiple $\mu$ ops) cannot reach this limit}.

@@

We explore the ability for the CPU to ``double speculate'', where a second
stalled indirect jump while the CPU is already speculating causes it to predict
the target and
speculate a second time. For instance, suppose a payload program truly jumps to
target A, but the CPU is mistrained by a trigger program that jumps to B, thus
causing the payload program to speculatively execute at B. At B, suppose there
is a second indirect jump, perhaps using the same register as the first jump
(which has still not resolved). If the trigger program jumps to C, the payload
program may speculate a second time and continue speculative execution at C.

We find that not all Intel CPUs support nested speculation. For
example, it appears Haswell chips do not speculate while already speculating,
but nonetheless support non-nested \speculake. Both Sandy Bridge (which
preceeded Haswell) and Kaby Lake (which followed Haswell) support nested
speculation. We find that when a CPU does support nested speculation, there
appears to be no limit to nested depth besides the speculative instruction
limit. We use a 16-deep nested speculation in Section~\ref{subsec:impl-aes}
to protect speculative decryption keys from reverse
engineering.

@@

We summarize our findings into a \emph{speculative primitive}, which allows our payload program to
speculatively (and covertly) perform on the order of 100 arbitrary
instructions while an accompanying trigger program is running, and communicate
a short (e.g. single byte) result to the real
world via a cache side channel. These speculative instructions are able to read
from any real-world memory or registers, but they cannot perform updates or
writes directly. To update memory, the speculative instructions must communicate
to the real world. We use a cache side channel to do so, but other side channels
compatible with Spectre could also be used~\cite{spectre1-1}.

We note a performance tradeoff between the size of communication (e.g. 4 bits vs 8
bits) and the time it takes the real world to recover the result from the side
channel. Using Flush+Reload~\cite{yarom2014flush+} as our cache side channel,
recovering the result requires accessing all elements in an array exponential in
the size of the result (e.g. $2^8$ array reads to recover an 8-bit result).
Therefore, there is a
performance advantage for keeping the size of the result small, and communicating
out small pieces of information that are aggregated by the real world.
Meanwhile, smaller channels introduce more overhead in recovering information. We
investigate this tradeoff in Section~\ref{subsec:impl-aes}, and find that 8~bits
is near optimal.
